### **4. Self-evaluation**

#### **Self-evaluation exam**

The first task for today is to complete an individual self-evaluating exam, where you will have **one hour** to answer some data analysis questions using R. 
This is **not for performance evaluation**, meaning that the correctness of your answers will not be evaluated, but your **application, participation, helping your colleagues** will be evaluated. 
It is mostly intended for you to identify your difficulties and consolidate knowledge.

1. For this self-evaluation exam, the `ualg.compbio` package must be installed (not needed if you are using the RStudio Server account):

```{r install_compbio_package, eval=FALSE, echo=TRUE}

# Make sure the `remotes` package is installed:
  # TRUE if installed, FALSE otherwise
"remotes" %in% rownames(installed.packages())  

# If FALSE, uncomment this line to install the package:
# install.packages("remotes")

# Now you can install the `ualg.compbio` package from GitHub:
remotes::install_github("instructr/ualg.compbio")

```

2. Run the `self_eval_exam` from the `ualg.compbio` package:

```{r run_self_eval, eval=FALSE, echo=TRUE}

learnr::run_tutorial("self_eval_exam", package = "ualg.compbio")

```

3. Please complete the exam. You have **60 minutes**.

---
